---
title: "Excess Rentals in TfL Bike Sharing"
author: "Harsh Tripathi"
summary: "A project focussed on Data manipulation and visualization."
tags:
- Data Manipulation
- Data Visualization
date: "2021-09-08T00:00:00Z"

# image: 
#   focal_point: Smart

output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
```

This data is about how many bikes were hired every single day in London. We will compare the monthly average with the overall average and plot the excess rentals for each month.

```{r, get_tfl_data}
#URL of dataset
url <- "https://data.london.gov.uk/download/number-bicycle-hires/ac29363e-e0cb-47cc-a97a-e216d900a6b0/tfl-daily-cycle-hires.xlsx"

#Download TFL data to temporary file
httr::GET(url, write_disk(bike.temp <- tempfile(fileext = ".xlsx")))

#Use read_excel to read it as dataframe
bike0 <- read_excel(bike.temp,
                   sheet = "Data",
                   range = cell_cols("A:B"))

#Change dates to get year, month, and week
bike <- bike0 %>% 
  clean_names() %>% 
  rename (bikes_hired = number_of_bicycle_hires) %>% 
  mutate (year = year(day),
          month = lubridate::month(day, label = TRUE),
          week = isoweek(day))
```

We calculate the expected number of rentals per month between 2016-2019 and compare how each month of 2020-2021 compares to the expected results.


```{r, monthly_average}
monthly_avg <- bike %>% 
  
  #Filtering for years 2016-19
  filter(year %in% c(2016,2017,2018,2019)) %>% 
  
  #Grouping by month
  group_by(month) %>%
  
  #Calculating mean bikes hired for each month
  summarise(
    mean_bikes_overall = mean(bikes_hired)
  )

monthly_avg
```

```{r, comparing_recent_data_with_avg}
bikes_monthly_mean <- bike %>% 
  
  #Filtering for recent years
  filter(year %in% c(2016,2017,2018,2019,2020,2021)) %>% 
  
  #Grouping by month and year
  group_by(month, year) %>% 
  
  #Calculating average bikes hired for each month
  summarize(
    mean_monthly_bikes = mean(bikes_hired)
  )

#Joining this data with overall average
bikes_with_avg <- left_join(bikes_monthly_mean, monthly_avg, on='month')

#Calculating excess rentals by comparing data with overall average
bikes_with_avg <- bikes_with_avg %>% 
  mutate(excess_rentals = mean_monthly_bikes - mean_bikes_overall)

#Defining an ordered list of months
month_labs <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

#Converting month to factor
bikes_with_avg$month <- factor(bikes_with_avg$month, label = month_labs, ordered = TRUE)
```


```{r, defining_variables_for_plot}
#Create variables "up" and "down" denoting the shifts from the average
bikes_with_avg_new<-bikes_with_avg %>%
  mutate(up = ifelse(excess_rentals > 0, as.numeric(excess_rentals), 0), down = ifelse(excess_rentals < 0, as.numeric(excess_rentals), 0)) %>% 
  arrange(year,month)

bikes_with_avg_new
```

```{r, plotting_excess_rentals, fig.width = 10}
#Line graph for mean bikes overall and mean monthly bikes
ggplot(bikes_with_avg_new, aes(x = month)) +
  geom_line(aes(group = 1, y = mean_bikes_overall), color = "blue") +
  geom_line(aes(group = 1, y = mean_monthly_bikes), color = "black", alpha = 0.6) +
  
  #Including lines and filling colours to show shift from mean
  geom_ribbon(aes(group = 1, ymin = mean_bikes_overall, ymax = mean_bikes_overall + up), fill = "#7DCD85", alpha = 0.4) +
  geom_ribbon(aes(group = 1, ymin = mean_bikes_overall, ymax = mean_bikes_overall + down),fill = "#CB454A",alpha = 0.4)+
  
  #Facet wrap by year
  facet_wrap(~year) +
  
  #Using black and white theme
  theme_bw() + 
  
  #Adding title and axes labels
  labs(
    title = "Monthly changes in TfL bike rentals",
    subtitle = "Change from monthly average shown in blue\nand calculated between 2016-2019 ",
    y = "Bike rentals",
    x = "",
    caption = "Source:Tfl, London Data Store"
    
  )
```

The blue line displays the overall monthly average of bikes hired (based on years 2016-2019). The black line displays the average monthly bikes hired in the given year. The filled area displays excess/deficit bikes as compared to the overall monthly average.


Now, we look at the percentage change from the expected level of weekly rentals. The two grey shaded rectangles correspond to Q2 (weeks 14-26) and Q4 (weeks 40-52).

```{r, data_manipulation_for_2nd_graph}
#Change certain dates (as some days of the week at the end or the start of the year are counted towards the wrong year)

bike1 <- bike %>% 
  
  #Converting month to numeric
  mutate(month=as.numeric(month)) %>%
  
  #Fixing the edge cases
  mutate(year = ifelse(month == 12 & week == 1, year + 1, year)) %>% 
  mutate(year = ifelse(month == 1 & week == 53, year-  1, year))

#Calculate weekly averages
weekly_avg <- bike1 %>%
  
  #Filtering for years 2016-19
  filter(year %in% c(2016,2017,2018,2019)) %>% 
  
  #Grouping by week
  group_by(week) %>%
  
  #Calculating average bikes hired weekly
  summarise(
    mean_weekly_overall = mean(bikes_hired)
  )

weekly_avg
```


```{r}
# #edit for week 53(some days are counted towards the wrong year)
# weekly_avg_53 <- bike1 %>% 
#   filter(year == 2015 & week==53) %>% 
#   group_by(week) %>% 
#   summarise(
#     mean_weekly_overall = mean(bikes_hired)
#   )
# 
# #combine the dataframes to create final dataframe
# weekly_avg_new<-full_join(weekly_avg,weekly_avg_53,on='week')
# weekly_avg_new
```




```{r, weekly_avg_for_recent_data}
bikes_weekly_mean <- bike1 %>%
  
  #Filtering for recent years
  filter(year %in% c(2016,2017,2018,2019,2020,2021)) %>%
  
  #Grouping by year and week
  group_by(year,week) %>%
  
  #Calculating average bikes hired for each week
  summarize(
    mean_weekly_bikes = mean(bikes_hired)
  )

bikes_weekly_mean
```



```{r, calculating_percentage_change}
#Joining recent average with overall average
bikes_week_avg <- left_join(bikes_weekly_mean, weekly_avg, on='week')

#Calculating excess weekly rentals
bikes_week_avg <- bikes_week_avg %>% 
  mutate(excess_weekly_rentals = mean_weekly_bikes - mean_weekly_overall)

#Defining variables for graph
bikes_week_avg_new <- bikes_week_avg %>%
  
  #Calculating excess weekly rentals
  mutate(excess_weekly_rentals_percent = (excess_weekly_rentals / mean_weekly_overall) * 100,
         
  #Defining up and down variables to determine fill
  up = ifelse(excess_weekly_rentals > 0, as.numeric(excess_weekly_rentals / mean_weekly_overall) * 100, 0),
  
  down = ifelse(excess_weekly_rentals < 0, as.numeric(excess_weekly_rentals / mean_weekly_overall) * 100, 0)) %>%
  
  #Sorting by year and week
  arrange(year,week)

bikes_week_avg_new
```

```{r, dropping_NAs}
bikes_week_avg_new <- bikes_week_avg_new %>% drop_na()
```

```{r, plotting_percent_change}, fig.width = 10}
#Line plot for excess weekly percent
ggplot(bikes_week_avg_new, aes(x = week)) + 
  geom_line(aes(group=1, y = excess_weekly_rentals_percent), color = "black", alpha = 0.3) +

  #Defining areas denoting shifts from the average
  geom_ribbon(aes(group = 1, ymin = 0, ymax = up), fill = "#7DCD85", alpha = 0.4) +
  geom_ribbon(aes(group = 1, ymin = 0, ymax = down), fill = "#CB454A", alpha = 0.4) +
  
  #Facet wrap by year
  facet_wrap(~year) +
  
  #Defining scales
  scale_y_continuous(breaks = c(-50, 0, 50,100), labels = function(x) paste0(x, "%"), limits = c(-50,100)) +
  scale_x_continuous(breaks = seq(0,53,by=13), limits = c(1,53)) +
  
  #Using black and white theme
  theme_bw() +
  
  #Using grey areas to shade 2nd and 4th quarters of the year
  annotate("rect", xmin=14, xmax=26, ymin=-Inf, ymax=Inf, alpha=0.4, fill="grey") +
  annotate("rect", xmin=40, xmax=52, ymin=-Inf, ymax=Inf, alpha=0.4, fill="grey") +
  
  #Including lines at axis
  geom_rug(data = bikes_week_avg_new[bikes_week_avg_new$up > 0,], sides = "b", colour = "#7DCD85") +
  geom_rug(data = bikes_week_avg_new[bikes_week_avg_new$down < 0, ], sides = "b", colour = "#CB454A") +
  
  #Setting title and axes labels
  labs(
    title = "Weekly changes in TfL bike rentals",
    subtitle = "% Change from weekly averages\ncalculated between 2016-2019 ",
    y = "",
    x = "Week",
    caption = "Source:Tfl, London Data Store")
```


This graph shows the weekly changes in Tfl bike rentals from the weekyl averages calculated on data between 2016-2019. The shaded region shows the percentage deviation of each week from the weekly average. The grey shaded regions highlight the 2nd and 4th quarter of every week.
